A first look at the GD-1 stellar stream with Gaia DR2

### 2018-04-28 paper/gd1.tex
Intro

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{PanSTARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\rm M_\odot}

@@

Dynamically cold stellar streams are formed from the tidal disruption of stellar
systems by the gravitational field of their host galaxy (e.g.,
\citealt{Johnston:1998}).
Over 30 candidate stellar streams have been discovered throughout the halo of
the Milky Way and have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
recent review \citealt{Grillmair:2006, Newberg:2016}).
Of the thin, likely globular-cluster-origin streams, only two have clear
progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).

Stellar streams are

\citealt{Bonaca:2018}

Dynamical modeling of stellar streams or models of mean stream tracks have been used

Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.

We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.

@@

Figure 1:
- PM cut (panel a)
- PM + CMD cut (panel b)
- SFD (panel c)??

>>>>

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{Pan-STARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\textrm{M}_\odot}
\newcommand{\kpc}{\textrm{kpc}}
\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
\newcommand{\article}{\textsl{Letter}}

@@

Dynamically cold stellar streams are formed from the tidal disruption of stellar
systems by the gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
Most stellar streams are found in the halos of galaxies and are therefore
important tracers of the large-scale distribution of dark matter in galactic
halos.

Long, thin stellar streams are also excellent laboratories for studying
small-scale structures in the mass distribution:
thin streams can remain coherent for tens of orbital periods, depending in
detail on the symmetries of the underlying galactic mass distribution (e.g.,
\citealt{Erkal:2016}).
Encounters between massive perturbers and stream stars imprint variations in the
phase-space density that can persist and remain visible for close to their total
survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
studying or ruling out the existence of small-scale dark matter sub-halos, as
would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

Over 30 candidate stellar streams have been discovered throughout the halo of
the Milky Way, largely thanks to large-area, multi-band photometric surveys
(especially the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
The Milky Way streams have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
also the recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority
have no obvious, bound progenitor systems;
of the thin, likely globular-cluster-origin streams, only two have clear
progenitors (NGC 5466 and Palomar 5).

The most prominent and longest (in angle) thin stream is the GD-1 stream, named
after its discoverers (\citealt{Grillmair:2006}).
GD-1 was found using a matched filter on photometric data from the SDSS and
spans over $60^\circ$ in Heliocentric sky coordinates, partially because of its
relative closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
Despite its proximity, no remnant progenitor cluster or stellar system has been
found associated with the stream, but its metallicity ($\feh \approx -1.4$;
\citealt{Koposov:2010}) and estimated stellar mass ($M \approx 2 \times
10^4~\msun$; \citealt{Koposov:2010}) are consistent with being a disrupted
globular cluster.

The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way halo.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates (\citealt{Bovy:2016}), it has been shown that the dark
matter distribution within its orbit is consistent with or nearly spherical.

Its long length --- and therefore its implied old dynamical age --- and its
particular orbit also make it a prime stream to search for interactions with
dark matter subhalos:
because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar, density variations in the stream are not expected from either
interactions with giant molecular clouds (\citealt{TODO}) or resonant
interactions with the bar (\citealt{Pearson:2017}).
It is particularly exciting, therefore, that deep photometry of the stream has
revealed interesting small-scale ``wiggles'' and density variations along a
$\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}).
, but color-magnitude only, bin stars]

In this \article, we present a the cleanest view of the GD-1 stream using
astrometric data from the \gaia\ mission data release 2 (\DR).
Combined with photometry from \pans\ ...
We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.

@@
